---
title: "Prison metrics"
author: "Chris Hoover et al"
date: "4/26/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(tidyverse)
```

```{r data}
# Data -----------------
dat           <- readRDS(here::here("data/derived/state_prisons_pop_cases_fin2021-03-18.rds"))

# NAs in county FIPS causing join to fill in NAs, so need to make sure all obs have proper county fips
dat_lookup_fips <- dat %>% filter(Date == "2020-04-01") %>% dplyr::select(Facility, County.FIPS)

dat <- dat %>% 
  left_join(dat_lookup_fips,
            by = "Facility",
            suffix = c(".x","")) %>% 
  dplyr::select(-County.FIPS.x)

CA_cases_pops <- readRDS(here::here("data/derived/CA_Cases_Pops2021-03-18.rds")) %>% 
  group_by(County) %>% 
  # Add measures of cumulative prevalence
  mutate(cum_cases_7day = zoo::rollsum(cases, k =7,
                                       align = "right", na.pad = T),
         prev_7day = cum_cases_7day/population,
         prev_7day_p100k = prev_7day*1e5,
         cum_cases_14day = zoo::rollsum(cases, k =14,
                                        align = "right", na.pad = T),
         prev_14day = cum_cases_14day/population,
         prev_14day_p100k = prev_14day*1e5) %>% 
  ungroup()

dat_case_pop_merge <- dat %>% 
  left_join(CA_cases_pops %>% mutate(County.FIPS = as.numeric(fips)),
            by = c("County.FIPS" = "County.FIPS",
                   "Date" = "date"),
            suffix = c(".x",""))
```

# Cases following introductions  

```{r ob_define_fig}
# define_outbreaks -----------------
# 14 day washout period      
outbreaks_df14day <- dat_case_pop_merge %>% 
  dplyr::select(-c(Residents.Recovered:New_Staff_Deaths, Residents.Tadmin:fips)) %>%
  group_by(Facility) %>% 
  mutate( # Identify outbreaks as new cases emerging following 14 days with no cases
    New_Residents_Confirmed_rmv_neg = if_else(Date == as.Date("2020-04-01"), 0, New_Residents_Confirmed_rmv_neg), #Remove NA on first obs day
    new_cases_14day = zoo::rollsum(New_Residents_Confirmed_rmv_neg, k = 14, 
                                   na.pad = T, align = "right"),
    new_cases_14day_lead1 = lead(new_cases_14day),
    outbreak_start = if_else(new_cases_14day == 0 & new_cases_14day_lead1 > 0, 1, 0),
    # Give each outbreak a unique identifier
    outbreak_num = cumsum(if_else(is.na(outbreak_start), 0, outbreak_start)) + outbreak_start*0,
    Facility_Outbreak = paste0(Facility, " Outbreak ", outbreak_num),
    plot7day_cases = if_else(new_cases_14day == 0, NA_real_, New_Residents_Confirmed_7day),
    # For censoring outbreaks with high prior case counts (high chance of effects of immunity)
    Cum_Residents_Confirmed_rmv_neg = cumsum(New_Residents_Confirmed_rmv_neg),
    Cum_attack_rate = Cum_Residents_Confirmed_rmv_neg/N0,
    Ns = N0 - Cum_Residents_Confirmed_rmv_neg
  ) %>% 
  ungroup()

# Plot time series with outbreaks delineated -----------------------
#Facilities with at least three outbreaks
g5_ob_facs <- outbreaks_df14day %>% 
  group_by(Facility) %>% 
  summarise(n_obs = max(outbreak_num, na.rm = T)) %>% 
  filter(n_obs >= 5) %>% 
  pull(Facility)

outbreaks_14d_wash_plot <- outbreaks_df14day %>% 
  filter(!is.na(outbreak_num),
         Facility %in% g5_ob_facs,
         outbreak_num > 0) %>% 
  mutate(Facility2 = str_remove(unlist(str_split(Facility, pattern = "\\("))[seq(2,nrow(.)*2, by=2)], pattern = "\\)")) %>% 
  ggplot() +
  geom_line(aes(x = Date, y = plot7day_cases, col = as.factor(outbreak_num))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 315, hjust = 0, size = 8),
        axis.title = element_text(size = 14),
        strip.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA)) +
  scale_y_continuous(trans = "log1p",
                     breaks = c(1,10,100,1000)) +
  scale_x_date(date_labels = "%b'%y") +
  facet_wrap(facets = "Facility2",
             nrow = 2, ncol = 5) +
  labs(y = "7-day average of incident resident cases",
       col = "Introduction")

outbreaks_14d_wash_plot

ggsave(here::here("Plots/Outbreaks_5min.png"),
       height = 4, width = 6, units = "in")
```

# Occupancies  
## All facilities  
```{r avg_pop_fig, fig.height=3, fig.width=5, fig.cap="Mean facility population size across all California state prison facilities. Shaded region indicates max and min while solid line represents the mean."}
pop_dat <- dat %>% 
  group_by(Date) %>% 
  summarise(avg_pop = mean(Nt),
            max_pop = max(Nt),
            min_pop = min(Nt))

avg_start <- pop_dat$avg_pop[which(pop_dat$Date == min(pop_dat$Date))]
avg_end <- pop_dat$avg_pop[which(pop_dat$Date == (max(pop_dat$Date)-5))]

pop_dat %>% 
  ggplot(aes(x = Date, y = avg_pop, ymin = max_pop, ymax = min_pop)) +
    geom_line() +
    geom_ribbon(alpha = 0.2) +
    theme_classic() +
    labs(x = "Date",
         y = "Facility populations")
```

Average occupancy across all prison facilities was `r round(avg_start)` as of `r min(pop_dat$Date)`, but had decreased to `r round(avg_end)` as of `r (max(pop_dat$Date)-5)`.

## Facilities shown in figure (with at least five introductions)  
```{r avg_pop_fig_g5_ob, fig.height=3, fig.width=5, fig.cap="Mean facility population size across California state prison facilities with at least 5 introductions. Shaded region indicates max and min while solid line represents the mean."}
pop_dat_g5_ob <- dat %>% 
  filter(Facility %in% g5_ob_facs) %>% 
  group_by(Date) %>% 
  summarise(avg_pop = mean(Nt),
            max_pop = max(Nt),
            min_pop = min(Nt))

avg_start_g5_ob <- pop_dat_g5_ob$avg_pop[which(pop_dat_g5_ob$Date == min(pop_dat_g5_ob$Date))]
avg_end_g5_ob <- pop_dat_g5_ob$avg_pop[which(pop_dat_g5_ob$Date == (max(pop_dat_g5_ob$Date)-5))]

pop_dat_g5_ob %>% 
  ggplot(aes(x = Date, y = avg_pop, ymin = max_pop, ymax = min_pop)) +
    geom_line() +
    geom_ribbon(alpha = 0.2) +
    theme_classic() +
    labs(x = "Date",
         y = "Facility populations")
```

Average occupancy across  prison facilities with at least five introductions was `r round(avg_start_g5_ob)` as of `r min(pop_dat_g5_ob$Date)`, but had decreased to `r round(avg_end_g5_ob)` as of `r (max(pop_dat_g5_ob$Date)-5)`.

# Introduction rate  
## All facilities  
```{r intros}
n_days <- length(unique(dat$Date))

n_facs <- length(unique(dat$Facility))

n_intros <- outbreaks_df14day %>% 
  filter(outbreak_num > 0) %>% 
  pull(Facility_Outbreak) %>% 
  unique() %>% 
  length()
```

Across `r n_days` days of observation of `r n_facs` facilities, there were `r n_intros` introductions, defined as a new resident case following a 14-day washout period with no resident cases, giving a rate of `r round(n_intros/n_facs/n_days, 3)` introductions per facility per day.

## Facilities shown in figure (with at least five introductions)  
```{r intros_g5_ob}
n_days_g5_ob <- length(unique(dat$Date[which(dat$Facility %in% g5_ob_facs)]))

n_facs_g5_ob <- length(unique(dat$Facility[which(dat$Facility %in% g5_ob_facs)]))

n_intros_g5_ob <- outbreaks_df14day %>% 
  filter(Facility %in% g5_ob_facs, outbreak_num > 0) %>% 
  pull(Facility_Outbreak) %>% 
  unique() %>% 
  length()
  
```

Across `r n_days_g5_ob` days of observation of `r n_facs_g5_ob` facilities, there were `r n_intros_g5_ob` introductions, defined as a new resident case following a 14-day washout period with no resident cases, giving a rate of `r round(n_intros_g5_ob/n_facs_g5_ob/n_days_g5_ob, 3)` introductions per facility per day.

# Outbreak rate  

## All facilities  
```{r ob_rate_all_facilities}
outbreak_size14day <- outbreaks_df14day %>% 
  filter(outbreak_num > 0) %>% 
  group_by(Facility_Outbreak) %>% 
  summarise(outbreak_size = sum(New_Residents_Confirmed_rmv_neg))

# Define upper cutoff for outbreak ----------------
upper_cutoff1 <- 10

outbreak_size14day_10uppercut <- outbreak_size14day %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff1, upper_cutoff1+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())


# Plot outbreak size distribution with upper cutoff-------------------
outbreak_size_distn <- outbreak_size14day_10uppercut %>% 
  ggplot(aes(x = outbreak_size_uppercut, y = n_ob)) +
    geom_col() +
    geom_col(data = outbreak_size14day_10uppercut %>% filter(outbreak_size_uppercut == (upper_cutoff1+1)),
             fill = "darkred") +
    theme_classic() +
    theme(axis.text = element_text(size = 14),
          axis.title = element_text(size = 16)) +
    scale_x_continuous(breaks = c(1:(upper_cutoff1+1)),
                       labels = c(as.character(c(1:(upper_cutoff1))),
                                  paste0(">", upper_cutoff1))) +
    labs(x = "Outbreak size (number of confirmed cases)",
         y = "Frequency",
         title = "Outbreak size distribution with 14-day washout")
    
outbreak_size_distn

ob_size1 <- outbreak_size14day_10uppercut %>% filter(outbreak_size_uppercut == 1) %>% pull(n_ob)

ob_size11 <- outbreak_size14day_10uppercut %>% filter(outbreak_size_uppercut == 11) %>% pull(n_ob)

```

## Facilities shown in figure (with at least five introductions)  

Of `r n_intros` introductions, `r ob_size1` (`r round(ob_size1/n_intros, 2)*100`%) resulted in no onward transmission and `r ob_size11` (`r round(ob_size11/n_intros, 2)*100`%) resulted in outbreaks of more than ten resident cases.

```{r ob_rate_g5obs_fac}
outbreak_size14day_g5ob <- outbreaks_df14day %>% 
  filter(Facility %in% g5_ob_facs, outbreak_num > 0) %>% 
  group_by(Facility_Outbreak) %>% 
  summarise(outbreak_size = sum(New_Residents_Confirmed_rmv_neg))


outbreak_size14day_10uppercut_g5ob <- outbreak_size14day_g5ob %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff1, upper_cutoff1+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())


# Plot outbreak size distribution with upper cutoff-------------------
outbreak_size_distn_g5ob <- outbreak_size14day_10uppercut_g5ob %>% 
  ggplot(aes(x = outbreak_size_uppercut, y = n_ob)) +
    geom_col() +
    geom_col(data = outbreak_size14day_10uppercut_g5ob %>% filter(outbreak_size_uppercut == (upper_cutoff1+1)),
             fill = "darkred") +
    theme_classic() +
    theme(axis.text = element_text(size = 14),
          axis.title = element_text(size = 16)) +
    scale_x_continuous(breaks = c(1:(upper_cutoff1+1)),
                       labels = c(as.character(c(1:(upper_cutoff1))),
                                  paste0(">", upper_cutoff1))) +
    labs(x = "Outbreak size (number of confirmed cases)",
         y = "Frequency",
         title = "Outbreak size distribution with 14-day washout",
         subtitle = "Among facilities with at least 5 introductions")
    
outbreak_size_distn_g5ob

g5_ob_size1 <- outbreak_size14day_10uppercut_g5ob %>% filter(outbreak_size_uppercut == 1) %>% pull(n_ob)

g5_ob_size11 <- outbreak_size14day_10uppercut_g5ob %>% filter(outbreak_size_uppercut == 11) %>% pull(n_ob)

```

Of `r n_intros_g5_ob` introductions, `r g5_ob_size1` (`r round(g5_ob_size1/n_intros_g5_ob, 2)*100`%) resulted in no onward transmission and `r g5_ob_size11` (`r round(g5_ob_size11/n_intros_g5_ob, 2)*100`%) resulted in outbreaks of more than ten resident cases.
