---
title: "Prisons Empirical MRD"
author: "Chris Hoover"
date: "2/18/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(zoo)
library(tidyverse)
```

## Methods  
COVID-19 data for all state prison facilities operated by the California Department of Corrections and Rehabilitation (CDCR) are reported in a CDCR [data dashboard](https://www.cdcr.ca.gov/covid19/population-status-tracking/). Machine readable timeseries of these daily reports were acquired from the [UCLA COVID Behind Bars](https://github.com/uclalawcovid19behindbars) project which gathers and organizes COVID 19 data from carceral facilities across the United States, including CDCR. Empirical analyses described below utilize data on cumulative active cases reported separately for residents and staff for each facility. Time series of incident cases were derived by taking the daily difference of reported cumulative cases and removing anomalies such as negative case counts.

### Definining outbreaks  
Outbreaks were defined in each facility as an increase in incident cases among residents following a wash out period in which no incident cases occurred. We chose a washout period of 10 days, approximately double the estimated serial interval for COVID 19 according to a recent systematic review [CITE](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7448781/#:~:text=The%20pooled%20estimate%20of%20serial,SD%20%3D%203.8)%20days%5D.). Wash out periods of 7 and 14 days were also used in sensitivity analyses.

### Estimating $R_{eff}$ from outbreak size distribution  



```{r getdata}
dat           <- readRDS(here::here("data/derived/state_prisons_pop_cases_fin2021-02-02.rds"))
CA_cases_pops <- readRDS(here::here("data/derived/CA_Cases_Pops2021-02-18.rds"))
```

```{r analysis_data}
dat_case_pop_merge <- dat %>% 
  left_join(CA_cases_pops %>% mutate(County.FIPS = as.numeric(fips)),
            by = c("County.FIPS" = "County.FIPS",
                   "Date" = "date"))

```

```{r fxs}
#Simulate clusters that occur after an introduction -----------------
#Outputs is a table theat lists the number of clusters of an individual size, and the numeber that lead to full blow outbreaks (last row of output)
#Inputs are:
#   the reproduciton number (R)
#   the dispersion parameter (k)
#   the size at which point an outbreak is deemed to occur (thresh)
#   the number of introductions to simulate (number)

sim_cluster <-function(R,k,thresh,number) {
  all_sizes <- tibble(i = 1:number) %>% group_by(i) %>% do({
    num_inf <- 1
    cluster_size <- 1
    while (num_inf > 0) {
      num_inf <- sum(rnbinom(num_inf,size = k, mu = R))
      cluster_size <- cluster_size + num_inf
      if(cluster_size > thresh) {
        cluster_size <- thresh + 1
        num_inf <- 0
      }
    }
    tibble(cluster_size = cluster_size)
  })
  all_sizes %>% group_by(cluster_size, .groups = "drop") %>% summarize(n = n())
}


# prob that i cases cause j cases ----------------
calc_r_ij <- function (i,j,r,k) {
  log_r_ij <- lgamma(j +k*i) - lgamma(j+1) - lgamma(k*i) + 
    k * i * log(k/(r+k)) +
    j * log(r/(r+k))
  r_ij<- exp(log_r_ij)
}

# Likelihood of data -----------------------
#
# thresh denotes the size cutoff that determines when a cluster becomes an outbreak.
# c_j_arr = probability of having a cluster of size j
calc_cluster_logL <- function(par, data, thresh) {
  R <- exp(par[1])
  k <- exp(par[2])
  
  cluster_size <- 1:thresh
  c_j_arr <- calc_r_ij(i=cluster_size,j=cluster_size-1,R,k)/cluster_size
  cluster_size[thresh+1] <- thresh + 1
  c_j_arr[thresh+1] <- 1-sum(c_j_arr)
  prob_arr <- tibble(cluster_size = cluster_size, prob = c_j_arr)
  
  # Truncate right tail
  data$cluster_size <- ifelse(data$cluster_size > thresh,thresh+1, data$cluster_size)

  #Do log L
  data <- left_join(data,prob_arr,by = 'cluster_size')
  nLL <- -sum(data$n * log(data$prob))
  
  return(nLL)
}

# Optimizer to find best fit Reff and k ----------------
reff_k_find <- function(INIT, DATA, THRESH){
  out_fit <- optim(par = INIT, 
                   fn = calc_cluster_logL, 
                   # Additional functions to cal_cluster_logL
                   data = DATA,
                   thresh = THRESH,
                   #method = "L-BFGS-B",
                   #control = list("trace" = 1)
                   hessian = TRUE)
  
  if(out_fit$convergence == 0){
    return(list("logPars" = out_fit$par,
                "SEs" = sqrt(diag(solve(out_fit$hessian)))))
  } else {
    warning("No convergence")
    return(list("logPars" = NA,
                "SEs" = NA))
  }
  
}

# Likelihood of data with fixed k -----------------------
#
# thresh denotes the size cutoff that determines when a cluster becomes an outbreak.
# c_j_arr = probability of having a cluster of size j
calc_cluster_logL_fix_k <- function(R, k, data, thresh) {
  
  cluster_size <- 1:thresh
  c_j_arr <- calc_r_ij(i=cluster_size,j=cluster_size-1,R,k)/cluster_size
  cluster_size[thresh+1] <- thresh + 1
  c_j_arr[thresh+1] <- 1-sum(c_j_arr)
  prob_arr <- tibble(cluster_size = cluster_size, prob = c_j_arr)
  
  # Truncate right tail
  data$cluster_size <- ifelse(data$cluster_size > thresh,thresh+1, data$cluster_size)

  #Do log L
  data <- left_join(data,prob_arr,by = 'cluster_size')
  nLL <- -sum(data$n * log(data$prob))
  
  return(nLL)
}
# Optimizer to find best fit Reff with fixed k
reff_find <- function(R_INIT, K_est, DATA, THRESH){
  out_fit <- optim(par = R_INIT, 
                   fn = calc_cluster_logL_fix_k, 
                   # Additional functions to cal_cluster_logL
                   k = K_est,
                   data = DATA,
                   thresh = THRESH,
                   #method = "L-BFGS-B",
                   #control = list("trace" = 1)
                   hessian = TRUE)
  
  if(out_fit$convergence == 0){
    return(list("logPars" = out_fit$par,
                "SEs" = sqrt(diag(solve(out_fit$hessian)))))
  } else {
    warning("No convergence")
    return(list("logPars" = NA,
                "SEs" = NA))
  }
  
}
```


```{r define_outbreaks}
# 7 day washout period 
outbreaks_df7day <- dat %>% 
  dplyr::select(c(Facility:Residents_Active)) %>% 
  group_by(Facility) %>% 
  mutate( # Identify outbreaks as new cases emerging following 7 days with no cases
    new_cases_7day = zoo::rollsum(New_Residents_Confirmed_rmv_neg, k = 7, 
                                   na.pad = T, align = "right"),
    new_cases_7day_lead1 = lead(new_cases_7day),
    outbreak_start = if_else(new_cases_7day == 0 & new_cases_7day_lead1 > 0, 1, 0),
    # Give each outbreak a unique identifier
    outbreak_num = cumsum(if_else(is.na(outbreak_start), 0, outbreak_start)) + outbreak_start*0,
    Facility_Outbreak = paste0(Facility, " Outbreak ", outbreak_num),
    # For plotting outbreaks 
    plot7day_cases = if_else(new_cases_7day == 0, NA_real_, New_Residents_Confirmed_7day)
  ) %>% 
  ungroup() %>% 
  filter(!is.na(outbreak_num) & outbreak_num > 0)



# 10 day washout period 
outbreaks_df10day <- dat %>% 
  dplyr::select(c(Facility:Residents_Active)) %>% 
  group_by(Facility) %>% 
  mutate( # Identify outbreaks as new cases emerging following 10 days with no cases
    new_cases_10day = zoo::rollsum(New_Residents_Confirmed_rmv_neg, k = 10, 
                                   na.pad = T, align = "right"),
    new_cases_10day_lead1 = lead(new_cases_10day),
    outbreak_start = if_else(new_cases_10day == 0 & new_cases_10day_lead1 > 0, 1, 0),
    # Give each outbreak a unique identifier
    outbreak_num = cumsum(if_else(is.na(outbreak_start), 0, outbreak_start)) + outbreak_start*0,
    Facility_Outbreak = paste0(Facility, " Outbreak ", outbreak_num),
    plot7day_cases = if_else(new_cases_10day == 0, NA_real_, New_Residents_Confirmed_7day)
  ) %>% 
  ungroup() %>% 
  filter(!is.na(outbreak_num) & outbreak_num > 0)



# 14 day washout period      
outbreaks_df14day <- dat %>% 
  dplyr::select(c(Facility:Residents_Active)) %>% 
  group_by(Facility) %>% 
  mutate( # Identify outbreaks as new cases emerging following 14 days with no cases
    new_cases_14day = zoo::rollsum(New_Residents_Confirmed_rmv_neg, k = 14, 
                                   na.pad = T, align = "right"),
    new_cases_14day_lead1 = lead(new_cases_14day),
    outbreak_start = if_else(new_cases_14day == 0 & new_cases_14day_lead1 > 0, 1, 0),
    # Give each outbreak a unique identifier
    outbreak_num = cumsum(if_else(is.na(outbreak_start), 0, outbreak_start)) + outbreak_start*0,
    Facility_Outbreak = paste0(Facility, " Outbreak ", outbreak_num),
    plot7day_cases = if_else(new_cases_14day == 0, NA_real_, New_Residents_Confirmed_7day)
  ) %>% 
  ungroup() %>% 
  filter(!is.na(outbreak_num) & outbreak_num > 0)
```

```{r plot_outbreak_ts, fig.height = 6, fig.width = 8}
# Plot time series with outbreaks delineated -----------------------
outbreaks_10d_wash_plot <- outbreaks_df10day %>% 
  mutate(Facility2 = str_remove(unlist(str_split(Facility, pattern = "\\("))[seq(2,nrow(.)*2, by=2)], pattern = "\\)")) %>% 
  ggplot() +
    geom_line(aes(x = Date, y = plot7day_cases, col = as.factor(outbreak_num))) +
    theme_classic() +
    theme(axis.text = element_text(size = 14),
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title = element_text(size = 16)) +
    scale_y_continuous(trans = "log1p",
                       breaks = c(0,10,100,1000)) +
    facet_wrap(facets = "Facility2",
               nrow = 5, ncol = 7) +
    labs(y = "7-day average of incident resident cases",
         col = "Facility\nOutbreak\nNumber")

outbreaks_10d_wash_plot
```

```{r plot_outbreak_size_distn}
# Plot raw outbreak size distribution -------------------
outbreak_size7day <- outbreaks_df7day %>% 
  group_by(Facility_Outbreak) %>% 
  summarise(outbreak_size = sum(New_Residents_Confirmed_rmv_neg))


outbreak_size10day <- outbreaks_df10day %>% 
  group_by(Facility_Outbreak) %>% 
  summarise(outbreak_size = sum(New_Residents_Confirmed_rmv_neg))


outbreak_size14day <- outbreaks_df14day %>% 
  group_by(Facility_Outbreak) %>% 
  summarise(outbreak_size = sum(New_Residents_Confirmed_rmv_neg))

# Define upper cutoff for outbreak ----------------
upper_cutoff1 <- 10

outbreak_size7day_10uppercut <- outbreak_size7day %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff1, upper_cutoff1+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())

outbreak_size10day_10uppercut <- outbreak_size10day  %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff1, upper_cutoff1+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())

outbreak_size14day_10uppercut <- outbreak_size14day %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff1, upper_cutoff1+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())


# Plot outbreak size distribution with upper cutoff-------------------
outbreak_size_distn <- outbreak_size10day_10uppercut %>% 
  ggplot(aes(x = outbreak_size_uppercut, y = n_ob)) +
    geom_col() +
    geom_col(data = outbreak_size10day_10uppercut %>% filter(outbreak_size_uppercut == (upper_cutoff1+1)),
             fill = "darkred") +
    theme_classic() +
    theme(axis.text = element_text(size = 14),
          axis.title = element_text(size = 16)) +
    scale_x_continuous(breaks = c(1:(upper_cutoff1+1)),
                       labels = c(as.character(c(1:(upper_cutoff1))),
                                  paste0(">", upper_cutoff1))) +
    labs(x = "Outbreak size (number of confirmed cases)",
         y = "Frequency",
         title = "Outbreak size distribution with 10-day washout")
    
outbreak_size_distn
```

```{r plot_outbreak_size_distn2, eval = FALSE, include = FALSE}
upper_cutoff2 <- 5

outbreak_size7day_5uppercut <- outbreak_size7day %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff2, upper_cutoff2+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())

outbreak_size10day_5uppercut <- outbreak_size10day  %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff2, upper_cutoff2+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())

outbreak_size14day_5uppercut <- outbreak_size14day %>% 
  mutate(outbreak_size_uppercut = if_else(outbreak_size > upper_cutoff2, upper_cutoff2+1, outbreak_size)) %>% 
  group_by(outbreak_size_uppercut) %>% 
  summarise(n_ob = n())

outbreak_size10day_5uppercut %>% 
  ggplot(aes(x = outbreak_size_uppercut, y = n_ob)) +
    geom_col() +
    geom_col(data = outbreak_size10day_5uppercut %>% filter(outbreak_size10day_5uppercut == (upper_cutoff2+1)),
             fill = "darkred") +
    theme_classic() +
    theme(axis.text = element_text(size = 14),
          axis.title = element_text(size = 16)) +
    scale_x_continuous(breaks = c(1:(upper_cutoff2+1)),
                       labels = c(as.character(c(1:(upper_cutoff2))),
                                  paste0(">", upper_cutoff2))) +
    labs(x = "Outbreak size (number of confirmed cases)",
         y = "Frequency",
         title = "Outbreak size distribution with 10-day washout")

```

```{r old_outbreak_dist_base_plots, eval = FALSE, include = FALSE}
par(mfrow = c(3,1))
hist(outbreak_size7day$outbreak_size, breaks = 30,
     xlab = "Outbreak Size", main = "Outbreak size dist'n with 7 day washout")
hist(outbreak_size10day$outbreak_size, breaks = 30,
     xlab = "Outbreak Size", main = "Outbreak size dist'n with 10 day washout")
hist(outbreak_size14day$outbreak_size, breaks = 30,
     xlab = "Outbreak Size", main = "Outbreak size dist'n with 14 day washout")


jpeg(here::here("Plots/outbreak_size_distns_collate_gt10.jpg"),
     width = 6, height = 6, units = "in", res = 300)
par(mfrow = c(3,1),
    mar = c(3,3,1,0.2),
    mgp = c(1.75,0.75,0))

hist(outbreak_size7day_uppercut$outbreak_size_uppercut, breaks = 30,
     xlab = "", main = "Outbreak size dist'n with 7 day washout", 
     xlim = c(1,10), ylim = c(0,60))
hist(outbreak_size10day_uppercut$outbreak_size_uppercut, breaks = 30,
     xlab = "", main = "Outbreak size dist'n with 10 day washout", 
     xlim = c(1,10), ylim = c(0,60))
hist(outbreak_size14day_uppercut$outbreak_size_uppercut, breaks = 30,
     xlab = "Outbreak Size", main = "Outbreak size dist'n with 14 day washout", 
     xlim = c(1,10), ylim = c(0,60))

dev.off()

```

```{r validate_reff_k, eval = FALSE, include = FALSE, warning = FALSE, message = FALSE}
set.seed(430)

test_reffs <- sapply(1:100, function(...){
  test_dat <- sim_cluster(3,0.1,10,length(outbreak_size10day$outbreak_size))
  
  test_fit <- reff_k_find(INIT = c(2,0.1),
                          DATA = test_dat,
                          THRESH = 10)
  
  exp(test_fit$logPars)
})

hist(test_reffs[1,], breaks = 30, main = "Distribution of Reff estimates") ; abline(v = 3, lty = 2, lwd = 1.5, col = 2)
hist(test_reffs[2,], breaks = 30, main = "Distribution of k estimates") ; abline(v = 0.1, lty = 2, lwd = 1.5, col = 2)

```

```{r est_Reff_k, warning=FALSE, message=FALSE, eval = FALSE, include = FALSE}
# Estimate Reff and k from outbreak size distributions
wash_thresh <- expand.grid("Wash" = c(7,10,14),
                           "thresh" = c(3,5,10))

Reff_k_ests <- t(apply(wash_thresh,1,function(x){
  wash <- x[1]
  thresh = x[2]
  
  if(wash == 7){
    df_use = outbreak_size7day
  } else if(wash == 10){
    df_use = outbreak_size10day
  } else if(wash == 14){
    df_use = outbreak_size14day
  } else {
    NULL
  }
  
  outbreak_uppercut <- df_use %>% 
    mutate(cluster_size = if_else(outbreak_size > thresh, thresh+1, outbreak_size)) %>% 
    group_by(cluster_size) %>% 
    summarise(n = n())

  Reff_k <- reff_k_find(INIT = log(c(2, 0.1)),
                        DATA = outbreak_uppercut, 
                        THRESH = thresh)
  
  return(c("Reff" = exp(Reff_k$logPars[1]),
           "Reff_SE" = exp(Reff_k$SEs[1]),
           "k" = exp(Reff_k$logPars[2]),
           "k_SE" = exp(Reff_k$SEs[2])))
  
}))

Reff_k_results <- cbind(wash_thresh, Reff_k_ests)

Reff_k_boots <- sapply(1:1000, function(...){
  outbreak_uppercut <- outbreak_size10day %>% 
    mutate(cluster_size = if_else(outbreak_size > 10, 10+1, outbreak_size)) %>% 
    slice(sample(1:nrow(.), size = nrow(.), replace = T)) %>% 
    group_by(cluster_size) %>% 
    summarise(n = n())
    

  Reff_k <- reff_k_find(INIT = log(c(2, 0.1)),
                        DATA = outbreak_uppercut, 
                        THRESH = thresh)
  
  return(exp(Reff_k$logPars))

})

hist(unlist(Reff_k_boots)[seq(1,length(Reff_k_boots)*2, by = 2)],
     breaks = 30)
```

```{r est_Reff_fixed_k, warning=FALSE, message=FALSE}
# Estimate Reff and k from outbreak size distributions
wash_thresh_k <- expand.grid("Wash" = c(7,10,14),
                           "thresh" = c(3,5,10),
                           "k" = c(0.1,0.3,1))

Reff_ests <- t(apply(wash_thresh_k,1,function(x){
  wash <- x[1]
  thresh = x[2]
  k = x[3]
  
  if(wash == 7){
    df_use = outbreak_size7day
  } else if(wash == 10){
    df_use = outbreak_size10day
  } else if(wash == 14){
    df_use = outbreak_size14day
  } else {
    NULL
  }
  
  outbreak_uppercut <- df_use %>% 
    mutate(cluster_size = if_else(outbreak_size > thresh, thresh+1, outbreak_size)) %>% 
    group_by(cluster_size) %>% 
    summarise(n = n())

  Reff_est <- reff_find(R_INIT = log(2),
                      K_est = k,
                      DATA = outbreak_uppercut, 
                      THRESH = thresh)
  
  return(c("Reff" = exp(Reff_est$logPars[1]),
           "Reff_SE" = exp(Reff_est$SEs[1])))
  
}))

Reff_results <- cbind(wash_thresh_k, Reff_ests)

Reff_boots <- sapply(1:10000, function(...){
  outbreak_uppercut <- outbreak_size10day %>%
    mutate(cluster_size = if_else(outbreak_size > 10, 10+1, outbreak_size)) %>%
    slice(sample(1:nrow(.), size = nrow(.), replace = T)) %>%
    group_by(cluster_size) %>%
    summarise(n = n())


  Reff_est <- reff_find(R_INIT = log(2),
                        K_est = 1.0,
                        DATA = outbreak_uppercut,
                        THRESH = 10)

  return(exp(Reff_est$logPars))

}, simplify = T)



hist(Reff_boots, breaks = 30,
     main = "", xlab = expression(R[eff]), cex.lab = 1.2)

reff_boots_df <- as.data.frame(Reff_boots)

reff_boots_plot <- ggplot(reff_boots_df, aes(x = Reff_boots)) +
    geom_histogram() +
    theme_classic() +
    theme(axis.text = element_text(size = 14),
          axis.title = element_text(size = 16))
    labs(x = expression(R[eff]~estimate),
         y = "Frequency")

c(quantile(Reff_boots, 0.25),
  median(Reff_boots),
  quantile(Reff_boots, 0.75))
```
